Supplementary Material

### 2025-05-26 manuscript/manuscript.qmd
several updates
- manuscript in alpha version
- still pending b-cell gating methodology for the papel to be ready
- added script for joint hierarchical bayesian model

**Title**: Different Cardiac Autonomic Trajectories in Response to Exercise-Induced Stress

**Authors**: Matías Castillo-Aguilar,^1^ ..., Cristian Núñez-Espinosa,^1,2^ \*

::: {custom-style="Afiliation"}
^1^ Department/Affiliation, University/Institution, City, State/Province, Country

^2^ Department/Affiliation, University/Institution, City, State/Province, Country
:::

@@

Cristian Núñez-Espinosa, Escuela de Medicina, Universidad de Magallanes (UMAG), Punta Arenas, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: [Full Address]. Phone: [Phone Number].

@@

**Objective**: Briefly state the primary aim(s) of your research investigation.

**Material and methods**: Briefly describe the overall research design, participants/samples, procedures, and data analysis methods.

**Results**: Summarize the key findings of your study, including relevant data and statistical results.

**Conclusion**: State the main conclusions drawn from your research and their significance.

**Keywords**: List 3-5 keywords relevant to your research topic.

@@

<!--- 
OPENING HOOK SUGGESTION: 
Start with 2-3 sentences that link a clinically relevant syndrome/condition to an unresolved mechanistic question. 
- Define the condition using 1-2 seminal citations. 
- Highlight its epidemiological or public health burden. 
- Introduce the core system/mechanism under investigation as the study's focus.
--->
[opening paragraph here...]

<!--- 
## 1. **Mechanistic Context & Knowledge Gaps**

### 1.1. Foundational Understanding

LAYER 1 SUGGESTION: 
Summarize established foundational knowledge about the system/mechanism. 
- Cite 2-3 key papers that define current consensus. 
- Mention widely used metrics/tools in the field (e.g., "[METRIC] measures [PHENOMENON]").
--->
[established knowledge summary...]

<!--- 
### 1.2. Emerging Contradictions & Unanswered Questions

LAYER 2 SUGGESTION: 
Highlight emerging contradictions or unanswered questions. Use contrastive language:
- "While [X] is well-documented, [Y] remains controversial..."
- "Recent studies suggest [Z], but fail to address [GAP]..."
--->
[gap identification...]

<!--- 
### 1.3. Methodological Limitations

LAYER 3 SUGGESTION: 
Critique methodological limitations of current approaches. 
- Explain why traditional tools/metrics are insufficient (e.g., "Static measures overlook [DYNAMIC PROCESS]"). 
- Introduce the need for innovative methods (e.g., "[NEW METHOD] better captures [COMPLEXITY]").
--->
[methodological critique...]

<!--- 
## 2. **Methodological Rationale & Hypothesis**

### 2.1. Justification of Research Paradigm

TOOL 1 SUGGESTION: 
Justify your ecologically/clinically relevant paradigm. 
- Compare it to lab-based alternatives (e.g., "[REAL-WORLD ASSAY] mimics [CONDITION] better than [TRADITIONAL ASSAY]"). 
- Cite validation studies if available.
--->
[paradigm justification...]

<!--- 
### 2.2. Innovative Analytical/Metric Approach

TOOL 2 SUGGESTION: 
Explain your novel analytical/metric approach. 
- Highlight how it addresses Layer 3 limitations (e.g., "[COMPOSITE INDEX] integrates [MULTI-SYSTEM SIGNALS]"). 
- Cite technical validation papers.
--->
[method explanation...]

<!--- 
### 2.3. Hypothesis Formulation

HYPOTHESIS SUGGESTION: 
Explicitly state how your approach resolves the gap. Use the formula:
"By combining [TOOL 1] with [TOOL 2], this study tests the hypothesis that [CONDITION] disrupts [MECHANISM] through [PATHWAY]."
--->
[hypothesis statement...]

<!--- 
## 3. **Anticipated Impact & Translation**

### 3.1. Mechanistic Impact

MECHANISTIC IMPACT SUGGESTION: 
Explain how findings could refine fundamental understanding. Example:
"Clarifying [X] may disentangle [Y] from [Z] in [SYSTEM], resolving long-standing debates about [CONTROVERSY]."
--->
[mechanistic impact...]

<!--- 
### 3.2. Clinical Translation

CLINICAL TRANSLATION SUGGESTION: 
Link mechanisms to actionable insights. Example:
"Identifying [BIOMARKER/PATHWAY] could guide [INTERVENTION] for [SUBGROUP], addressing [UNMET CLINICAL NEED]."
--->
[clinical translation...]

<!--- 
### 3.3. Paradigm Shift in Methodology

PARADIGM SHIFT SUGGESTION: 
Highlight broader methodological implications. Example:
"This [METHODOLOGY] bridges [DISCIPLINE 1] and [DISCIPLINE 2], enabling future studies of [COMPLEX PHENOMENA]."
--->
[paradigm shift...]

<!--- 
## 4. **Additional Considerations**

### 4.1. Rhetorical Devices & Integration

RHETORICAL DEVICES SUGGESTION: 
- Use citation clusters thematically (e.g., group mechanistic studies separately from clinical papers). 
- Employ contrastive syntax to emphasize gaps ("Although...", "However..."). 
- Connect methods to hypotheses with bridging language ("To resolve this, we...").
--->
[narrative integration and rhetorical devices...]

<!--- 
Final Checklist for Blueprint Consistency

FINAL CHECKLIST: 
- [ ] Hook ties condition to mechanism in ≤3 sentences 
- [ ] Gap narrative progresses: Consensus → Contradictions → Method limitations 
- [ ] Hypothesis explicitly links tools to the gap 
- [ ] Impact addresses both basic science and translation
--->

@@

### Overall Study Framework  
<!---  
Describe the general design of your study (e.g., experimental, observational, clinical trial, cohort study, cross-sectional study, case-control study).  
Provide context for why this study type is most appropriate for addressing your research question.  
--->
[Your overall design description...]

### Structural Specifics and Justification  
<!---  
Detail the specific structure of the study (e.g., randomized, placebo-controlled, blind, longitudinal, cross-sectional).  
Explain the rationale behind choosing this structure, emphasizing its advantages over alternatives.  
Include any necessary visual aids such as a flowchart to illustrate the study design (optional but recommended).  
--->
[Your detailed design and justification...]

@@

### Recruitment and Sampling Strategy  
<!---  
Describe the methods used for recruitment or sample acquisition (e.g., clinics, community outreach, database screening).  
Outline the process for selecting participants to ensure a representative and unbiased sample.  
--->
[Your recruitment details...]

### Inclusion and Exclusion Criteria  
<!---  
Specify clear inclusion and exclusion criteria for participants or samples to ensure the study’s internal validity.  
Discuss any stratification or subgroup definitions if applicable.  
--->
[Your inclusion/exclusion criteria...]

### Sample Size and Demographics  
<!---  
Provide the sample size and justification for its adequacy (e.g., through power analysis).  
Describe key demographic characteristics (age, sex, ethnicity, etc.) and any variability considerations.  
--->
[Your sample size rationale and demographic details...]

### Ethical Considerations  
<!---  
Document the ethical aspects of the study, including:  
- Institutional review board (IRB) or ethics committee approval.  
- Informed consent procedures.  
- Compliance with animal welfare practices (if applicable).  
--->
[Your ethical procedures...]

@@

### Experimental Protocols and Interventions  
<!---  
Provide a comprehensive description of all experimental protocols, treatments, interventions, and control conditions.  
Detail key variables such as dosage, treatment duration, frequency, and administration method where relevant.  
--->
[Your experimental procedures...]

### Standardization and Instrumentation  
<!---  
Explain the instruments or equipment used for measurements, including calibration and standardization procedures to ensure reproducibility.  
Include details regarding the setup, maintenance, and quality assurance of these instruments.  
--->
[Your instrumentation details...]

@@

### Assessment Strategy  
<!---  
Outline the assessments conducted as part of the study.  
Explain how each assessment aligns with the study objectives and outcomes of interest.  
--->
[Your overall assessment strategy...]

### Assessment Protocols

#### Assessment 1

<!---  
Describe the specific variable(s) being measured in the first assessment.  
Detail the instrument(s) or techniques used, along with any calibration or validation steps.  
Include quality control measures to ensure data reliability.  
--->
[Detailed description of Assessment 1.]

#### Assessment 2

<!---  
Elaborate on the second assessment using the same structured approach:  
- Variable(s) measured  
- Instrumentation and methodology  
- Quality control and validation procedures  
--->
[Detailed description of Assessment 2.]

#### Assessment 3

<!---  
Provide a similar detailed protocol for a third assessment if applicable.  
--->
[Detailed description of Assessment 3.]

@@

<!---  
### 5.1. Data Analysis Methodology

Describe the overall statistical methods used to analyze the data.  
List the specific statistical tests applied (e.g., Student's t-test, ANOVA, linear regression, etc.).  
Outline how data distributions were assessed and any transformations applied.  
--->
[Your detailed data analysis methodology...]

<!---  
### 5.2. Software and Significance Criteria

Specify the statistical software and programming languages used (e.g., R, SPSS, SAS).  
Indicate the level of statistical significance (e.g., p < 0.05) and any adjustments for multiple comparisons if applicable.  
--->
[Your software and significance details...]

<!---  
### 5.3. Handling Confounders and Robustness Checks

Discuss the methods used to control for potential confounding variables.  
Include any sensitivity analyses or data robustness checks implemented to validate the results.  
--->
[Your methods to control confounders...]

<!---  
### 5.4. Data Quality Assurance

Outline the quality control procedures and verification methods for ensuring data integrity.  
This may include repeated measures, inter-rater reliability assessments, or external validation steps.  
--->
[Your quality assurance protocols...]

@@

@@

[bla bla bla].

@@

[bla bla bla].

@@

[bla bla bla].

@@

[bla bla bla].

@@

[bla bla bla].

@@

[bla bla bla].

@@

[bla bla bla].

>>>>

**Title**: Immune-Derived Signatures of RR-Interval Dynamics: Predicting Autonomic Responses from CD21/CD11c B-Cell Subsets

**Authors**: Matías Castillo-Aguilar,^1,2^ ..., Cristian Núñez-Espinosa,^1,2^ \*

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente y de Investigación (CADI-UMAG), Punta Arenas, Chile.

^2^ Escuela de Medicina, Universidad de Magallanes (UMAG), Punta Arenas, Chile
:::

@@

Cristian Núñez-Espinosa, Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

@@

**Objective**: This study aimed to investigate how specific B-cell subpopulations, CD21^+^/CD11c^+^ and CD21^-^/CD11c^-^, influence autonomic nervous system (ANS) dynamics, as measured by heart rate variability (HRV), in response to acute exercise in older adults.

**Material and Methods**: Using a cross-sectional design, 81 community-dwelling older adults (mean age 70.7 +/- 5.8 years) underwent immune profiling to quantify total B cells and four CD21/CD11c phenotypes. Continuous R-R interval (RRi) data were recorded at rest, during a Two-Minute Step Test (TMST), and over a five-minute recovery period. A coupled-logistic RRi-vs-time model yielded seven kinetic parameters per participant. Individual parameter estimates were regressed on standardized immune predictors using multivariate Bayesian models, adjusting for age and sex.

**Results**: Higher counts of CD21^+^/CD11c^+^ B cells were associated with elevated baseline RRi (resting vagal tone), a moderated exercise-induced RRi drop, and slower post-exercise recovery (posterior probability > 95%). Conversely, greater CD21^-^/CD11c^- B-cell counts related with lower resting RRi, a pronounced sympathetic-driven RRi decrease during exercise, and more rapid vagal reactivation during recovery (posterior probability > 90%).

**Conclusion**: CD21^+^/CD11c^+^ and CD21^-^/CD11c^-^ B-cell subsets exhibit divergent influences on ANS responsiveness to acute exercise in older adults. An integrative neuro-immunological approach may identify biomarkers of resilience or fragility, potentially informing personalized interventions to optimize cardiovascular health in aging.

**Keywords**: B Cells; Heart Rate Variability; Autonomic Nervous System; Aging; Exercise.

@@

The intricate interplay between the nervous and immune systems, often referred to as the neuro-immunological axis, is fundamental for maintaining physiological homeostasis and providing neuroprotection [@bonaz2017vagus]. Within this axis, the autonomic nervous system (ANS), with its sympathetic (SNS) and parasympathetic (PNS) branches, plays a pivotal role in bidirectional communication. Through the release of neurotransmitters such as norepinephrine (NE) and acetylcholine (ACh), the ANS modulates cardiovascular, metabolic, and immune functions. NE, released by sympathetic efferents, acts via $\alpha$- and $\beta$-adrenergic receptors to prepare the organism for action by increasing heart rate and vascular tone [@benarroch2020physiology]. Conversely, ACh, released by the vagus nerve, slows cardiac rhythm through muscarinic and nicotinic receptors, promoting energy conservation and anti-inflammatory effects [@benarroch2020physiology]. Immune cells express these receptor types, rendering them functional targets of autonomic regulation. Moreover, B and T lymphocytes can synthesize ACh via choline acetyltransferase, contributing to a non-neuronal cholinergic system within lymphoid tissues [@halder2021cholinergic].

The bidirectional nature of this axis implies that the ANS can modulate immune cell differentiation, cytokine production, and antibody responses, while in turn, immune mediators such as cytokines influence autonomic tone [@jacobson2021intestinal]. Among autonomic pathways, the vagus nerve plays a key role in limiting inflammation via the cholinergic anti-inflammatory reflex, which suppresses peripheral cytokine release, particularly tumor necrosis factor alpha (TNF-$\alpha$), in response to inflammatory stimuli [@tracey2002inflammatory; @bonaz2017vagus]. Disruption of this balance contributes to pathological processes including cardiovascular, metabolic, and neurodegenerative diseases [@giunta2024autonomic].

Aging alters both arms of this axis. On the immune side, immunosenescence manifests as diminished antigenic responsiveness, reduced naive lymphocyte pools, and an increased burden of memory and senescent immune cells, leading to a chronic low-grade inflammatory state, "inflammaging", marked by elevated circulating cytokines even in the absence of acute infection [@barrera2017inmunosenescencia; @frasca2023obesity; @giunta2024autonomic]. In parallel, the autonomic profile of older adults tends to shift toward sympathetic predominance and diminished parasympathetic tone, blunting the anti-inflammatory potential of vagal activity [@giunta2024autonomic]. These age-associated shifts not only increase susceptibility to chronic inflammatory conditions but also diminish physiological adaptability to external stressors by impairing the balanced responsiveness of the neuro-immunological axis.

Physical activity is a potent modulator of this dysregulation. Habitual exercise enhances vagal tone, increases heart rate variability (HRV), and reduces systemic inflammation [@olivieri2024heart]. From an immunological perspective, regular activity is associated with a more balanced T and B cell phenotype, including higher proportions of naive lymphocytes and reduced senescent subsets [@weyh2020physical; @tylutka2021lifestyle]. Exercise can thus be conceptualized as a physiological probe of neuroimmune flexibility, where the dynamic response of autonomic and immune markers reveals latent regulatory capacity.

In this context, B lymphocytes warrant particular attention. Beyond their role in antibody production, B cells function as antigen-presenting cells and cytokine producers, exerting both pro- and anti-inflammatory effects [@wang2020b]. Recent studies indicate that B cells also influence neural function and may contribute to neuroinflammatory conditions [@engler2024b]. With aging, the B cell compartment undergoes phenotypic remodeling, characterized by a decline in naive cells and expansion of memory and atypical populations. Among these, CD11c^+^ B cells, often termed age-associated B cells (ABCs), are enriched for pro-inflammatory and autoimmune signatures [@golinski2020cd11c]. They express transcriptional profiles associated with antigen experience and differentiation, including heightened expression of T-bet and Blimp-1. The co-expression of CD21, a complement receptor, provides further phenotypic granularity. CD21 is typically expressed on follicular B cells, but chronic immune stimulation can lead to its downregulation, indicating an exhausted or anergic state [@del2020rheumatoid].

Understanding the functional implications of these B cell subtypes in older adults may offer new insights into the coordination between immune aging and autonomic adaptability. During exercise, the ANS orchestrates a coordinated response involving sympathetic activation during exertion and parasympathetic reactivation during recovery. These changes are reflected in HRV, a time-domain measure of RR interval variability. High vagal tone is associated with greater HRV at rest and more rapid recovery post-exertion, whereas sympathetic dominance produces reduced HRV and delayed recovery [@castillo2023validity; @castillo2025enhancing].

Recent work has introduced parametric models to quantify individual differences in autonomic dynamics during and after exercise. These models extract parameters such as baseline HRV and recovery slope, which are thought to index autonomic flexibility and resilience [@castillo2025enhancing]. However, the immunological correlates of these parameters, particularly in older adults, remain poorly defined.

This study investigates whether baseline proportions of specific B cell subtypes, particularly CD21^+^/CD11c^+^ and CD21^-^/CD11c^-^ populations, are associated with interindividual differences in autonomic trajectories following submaximal exercise in older adults. By integrating immunophenotyping with HRV-based modeling of autonomic responses, we aim to elucidate how immune senescence interfaces with autonomic regulation. This may inform the identification of predictive biomarkers of physiological resilience and guide interventions targeting healthy aging.

@@

To explore the influence of CD21^+^/CD11c^+^ and CD21^-^/CD11c^-^ B cell subpopulations on autonomic dynamics, this study employed an observational, correlational, and cross-sectional design. This cross-sectional approach involved collecting data at a single time point, which enabled the examination of associations between the variables of interest.

Prior to any data acquisition, all participants received a comprehensive explanation of the study's aims, procedures, and potential outcomes. In adherence to ethical standards and to ensure respect for individual autonomy, informed consent was obtained from each participant.

## Setting

This study took place at the Centro Asistencial Docente e Investigacion (CADI-UMAG), an academic healthcare and research center associated with the University of Magallanes, in Punta Arenas, Chile. To minimize the impact of circadian rhythms on physiological measurements, all assessments were conducted in a controlled environment between 9:00 and 11:00 a.m.

The evaluation room was maintained at a consistent temperature of 20°C to ensure participant comfort and standardize testing conditions, thereby reducing potential influences of thermoregulation on autonomic responses. Consistent illumination was provided by artificial white lighting to prevent variations in ambient light that could affect visual or cognitive factors during the assessments. Furthermore, all evaluations were carried out in a private and quiet setting to minimize external distractions and enhance the reliability of the collected data.

@@

Participants were recruited from the local community through advertisements and outreach. A total of 81 older adults participated in this study, comprising 56 women and 25 men, with a mean age of 70.7 years (SD = 5.8, range 61-89). Inclusion criteria for participation were: (i) being 60 years or older at the time of enrollment; (ii) permanent residency in the Magallanes and Chilean Antarctic region, ensuring a relatively homogeneous population exposed to similar environmental and socioeconomic factors; (iii) achieving a score above 60% on the Karnofsky Performance Status scale, a common measure of functional capacity, indicating sufficient autonomy to complete the study assessments [@schag1984karnofsky]; and (iv) no prior diagnosis of conditions that could confound autonomic or cardiovascular function, such as diabetic neuropathy, pacemaker implantation, clinical depression, cognitive impairment, motor disability, or dementia.

Exclusion criteria were implemented to minimize factors that could confound autonomic measurements. Individuals were excluded if they: (i) were using beta-blockers during the study, as these medications can significantly alter autonomic and cardiovascular responses; (ii) had consumed any stimulant substances, including caffeine or sympathomimetic medications, within 12 hours prior to cardiac assessment; or (iii) had any degree of motor impairment restricting independent movement that could interfere with study procedures. Importantly, no recruited participants met these exclusion criteria.

The data for this study originated from a cohort involved in FONDECYT Project No. 11220116, funded by the Chilean National Agency for Research and Development (ANID). Ethical approval was granted by the Ethics Committee of the University of Chile (ACTA No. 029 - 18/05/2022) and the Ethics Committee of the University of Magallanes (No. 008/SH/2022).

@@

Participants attended a single study visit after a 12-hour fast and abstaining from strenuous exercise and alcohol. Upon arrival, they provided informed consent and underwent a brief health screening to re-verify eligibility based on the inclusion/exclusion criteria, including blood pressure assessment (SBP <140 mmHg and DBP <90 mmHg).

Sociodemographic data, including name, age, sex, and any existing chronic medical conditions, were collected via a structured interview. This information was used to characterize the study sample and identify potential confounding factors.

A 4 mL peripheral venous blood sample was collected in EDTA tubes by a trained nurse. These samples were gently homogenized and kept at room temperature for subsequent immunophenotypic analysis, being processed within two hours of collection.

Following a minimum 10-minute rest period in a seated position, participants were fitted with a Polar H10 heart rate monitor for continuous R-R interval recording and an Omron Hem-7142 monitor for blood pressure measurements. After baseline physiological measurements (HRV segment at rest, BP), participants performed the Two-Minute Step Test (TMST) as an acute physiological stressor. R-R intervals were recorded continuously during the 2-minute test and for a 5-minute recovery period immediately afterward. Blood pressure was measured again immediately after completion of the TMST to ensure participant safety.

@@

### Two-Minute Step Test (TMST)

The Two-Minute Step Test (TMST), a standardized protocol from the Senior Fitness Test battery [@rikli1999functional], was used as an acute, exercise-induced physiological stressor to challenge cardiac autonomic control. While also providing an index of lower body endurance and functional mobility, the TMST's primary role in this study was to elicit a controlled, submaximal physiological perturbation suitable for assessing dynamic autonomic responses in older adults. Participants were instructed to march in place for two minutes, lifting their knees to a midpoint between the patella and the iliac crest, guided by verbal cues. The total number of steps completed served as a measure of exercise volume and functional performance. A trained kinesiologist counted the steps, and participants received verbal encouragement to maintain consistent effort and proper form throughout the two-minute period. Rest was allowed if needed, with participants encouraged to resume the test as soon as possible.

During the TMST, R-R intervals were continuously recorded using Polar H10 heart rate monitors to assess HRV dynamics before, during, and after exercise. Additionally, blood pressure (BP) was measured immediately before and after the test to evaluate safety, physiological adaptations, and recovery using the Omron Hem-7142 monitor.

The TMST was performed after a 10-minute rest period, followed by a 5-minute recovery period (12 minutes total). The total number of steps completed in two minutes was recorded as the outcome measure for functional capacity, while the physiological stress induced by this standardized exertion served as the basis for assessing dynamic autonomic responses.

### Immune Profiling and B Cell Phenotyping

[...]

@@

We employed a fully Bayesian modeling framework to characterize how cardiac autonomic modulation responds to exercise in the presence of multiple confounding influences. Bayesian inference was preferred over traditional frequentist methods because it provides complete posterior distributions for all parameters, thereby allowing thorough quantification of uncertainty and probabilistic interpretation via credible intervals [@van2021bayesian]. By incorporating prior information, either drawn from existing literature or specified as weakly regularizing distributions, we constrained implausible parameter values, mitigated the impact of outliers, and improved convergence during model fitting.

For descriptive statistics, continuous variables are summarized as mean +/- standard deviation (M ± SD), while categorical variables are reported using absolute counts (n) and relative frequencies (%).

### RRi-vs-Time Model

To model the RR interval (RRi) across a complete rest-exercise-recovery protocol, we implemented the coupled-logistic RRi-vs-time model proposed by Castillo-Aguilar et al. (2025) [@Enhancing2025Castillo]. Under this specification, each observed RRi is assumed to follow a normal distribution centered on a time-dependent function. This data generation process is depicted in @eq-data-generation-process.

$$
\mathrm{RRi} \sim \mathcal{N}(f(t \mid \theta_k),\,\sigma^2)
$${#eq-data-generation-process}

Here, $\sigma^2$ represents the variance term, and the function $f(t \mid \theta_k)$ takes the form described in @eq-mca-model.

$$
f(t \mid \theta_k) = \alpha + \underbrace{\frac{\beta}{1 + e^{\lambda (t - \tau)}}}_{\mathrm{Drop\,Phase}} - \underbrace{\frac{c \beta}{1 + e^{\phi (t - \tau - \delta)}}}_{\mathrm{Recovery\,Phase}}
$${#eq-mca-model}

The parameter vector $\theta_k = \{\alpha,\beta,c,\lambda,\phi,\tau,\delta\}$ governs the shape of the RRi trajectory over time. Specifically, the parameters $\alpha$, $\beta$, and $c$ are magnitude parameters: $\beta$ controls the initial drop in RRi during exercise, $c\beta$ controls the subsequent recovery magnitude, and $\alpha$ serves as the baseline RRi level. The parameters $\lambda$ and $\phi$ are rate parameters that determine the steepness (i.e., the speed) of the drop and recovery transitions, respectively. Finally, the timing parameters $\tau$ and $\delta$ indicate when the drop phase begins and how the recovery phase onset is shifted relative to $\tau$.

### Two-Step Modeling Strategy

Our modeling proceeded in two distinct stages. First, for each participant, we obtained point estimates of the RRi-vs-time parameters (i.e., $\alpha,\beta,c,\lambda,\phi,\tau,\delta$) by fitting the RRi-vs-Time model using the R package *CardioCurveR* (v1.0.0). This package implements a box-constrained quasi-Newton algorithm with a Huber loss function for robust parameter estimation [@cardiocurverPackage].

Second, after deriving individual-level point estimates $\widehat{\theta}_{ik}$ for each subject $i$ and each curve-parameter $k$, we specified two multivariate Bayesian regression models in which each component $\widehat{\theta}_{ik}$ served as a response. Concretely, for the $k$th parameter of subject $i$ we let the conditional mean

$$
\mu_{k,i} = \beta_{k,0} + \sum_{p=1}^{6} \beta_{k,p} x_{p,i}
$${#eq-second-model}

where $x_{i1},\dots,x_{i6}$ denote the six centered and scaled immune-cell predictors (absolute lymphocyte count, total B-cell count, and the four B-cell phenotypes by CD21/CD11c combinations). In the first model, we regressed $\widehat{\theta}_{ik}$ on those six immune predictors via @eq-second-model. In the second model, we extended @eq-second-model by adding two additional terms, $\beta_{k,\mathrm{age}} (\mathrm{age}_i)$ and $\beta_{k,\mathrm{sex}} (\mathrm{sex}_i)$, to account for confounding by age and sex when modeling RRi kinetics. Prior to fitting either multivariate model, all continuous predictors (immune counts and age) were centered to their mean and scaled by their standard deviation, and sex was coded as 0/1. We specified $\mathcal{N}(0,3)$ priors on every main-effect coefficient $\beta_{k}$ to exert a weakly regularizing influence and to mitigate the impact of outliers.

### Model Fitting and Inference

Bayesian estimation employed the No-U-Turn Sampler (a variant of Hamiltonian Monte Carlo) as implemented in the *brms* (v2.22.0) and *rstan* (v2.32.7) packages in R [@rlanguage]. For each multivariate model, five Markov chains were run, each with 2000 warm-up iterations followed by another 2000 sampling iterations, resulting in 10000 post-warmup samples per parameter.

Inference followed the SEXIT (Sequential Effect eXistence and Significance Testing) framework [@makowski2019indices]. For each estimated parameter and its corresponding posterior distribution, we report the posterior median and its 95% highest-density credible interval (HDI). We also present the probability of direction (pd) as a quantitative measure of effect existence. Practical significance (ps) is indexed by the proportion of the posterior mass falling outside a region of practical equivalence (ROPE). The ROPE was defined as ±0.1 times the standard deviation of the response variable, and to ensure consistency all predictors were standardized prior to modeling [@makowski2019indices].

### Model Diagnostics

To verify sampling convergence and stability, we checked that the potential scale reduction factor ($\hat{R}$) for every parameter was below 1.01 and that the effective sample size exceeded 1,000. We also inspected trace plots visually to confirm adequate chain mixing and performed posterior predictive checks to ensure the model's predicted RRi distributions aligned with the observed data. All analyses were fully implemented in R [@rlanguage].

@@

```{r}
#| include: false

## Load libraries
library(brms)
library(data.table)
library(gtsummary)

## Load auxiliary functions
source("../R/_functions.R")

normalize_text <- function(x) {
    x <- stringi::stri_trans_general(x, "latin-ascii")
    clipr::write_clip(x)
    return(x)
}

## Load data objects
data <- readRDS("../models/m_data.RDS") # Data used in the models
model <- readRDS("../models/m_mod_2.RDS") # Adjusted model
```

@@

The samples consisted of 81 subjects with 70.6 ± 5.8 years old (age range, 61 to 89 years old), with 25 (30.9%) males and 56 (69.1%) females. The rest of the characteristics of the collected sample are presented in [table 1](#tab-1).

[**Table 1**]{#tab-1}. Sample characteristics in sociodemographical, immunological variables.

```{r}
#| echo: false

data[, sex := factor(sex, levels = c("F", "M"), labels = c("Females", "Males"))]
tbl_data <- data[, c(15:16,9:14)]
tbl_data |> 
  tbl_summary(by = sex,
            missing = "no",
            statistic = list(
              all_continuous() ~ "{mean} ± {sd}"
            ),
            label = list(
              age ~ "Age (years old)",
              abc_linfocitos_total ~ "Total lymphocytes (n)",
              abc_linfocitosb_total ~ "Total B cells (n)",
              cd21_m_cd11c_p ~ "CD21- CD11c+ B cells (n)",
              cd21_p_cd11c_p ~ "CD21+ CD11c+ B cells (n)",
              cd21_p_cd11c_m ~ "CD21+ CD11c- B cells (n)",
              cd21_m_cd11c_m ~ "CD21- CD11c- B cells (n)"
            )) |> 
  add_overall() |> 
  add_difference(test = list(all_continuous() ~ "smd")) |> 
  modify_spanning_header(all_stat_cols(stat_0 = F) ~ "**Sex**")
```

@@

Exercise elicits a U-shaped form on RRi during a rest-exercise-recovery protocol. This shape is controlled by the model parameters depicted in [Figure 2](#fig-2). <!--- Figure 1 will be gating strategy for B cell CD21 and CD11c --->

![](../figures/hand-designed/rri-curve.pdf){width=100%}

[**Figure 2**]{#fig-2}. Cardiac autonomic trajectories given by RRi dynamics in response to exercise as described by the RRi-vs-time model. This image also depicts the rest-exercise-recovery protocol used to induce the autonomic changes in response to physical exertion. The parameters $\alpha,\beta,c$ control the magnitude of change, $\lambda,\phi$ control the steepnees or speed of change, and $\tau,\delta$ control de timing of these changes in RRi dynamics.

<!--- ## See output from "R/rri-characteristics.R"
   Parameters Estimate    SE               CI
       <fctr>    <num> <num>           <char>
1:      alpha   876.89 15.29 [845.45, 906.55]
2:       beta   386.64 17.99 [350.97, 420.27]
3:          c     0.90  0.02     [0.86, 0.93]
4:     lambda     2.96  0.16     [2.63, 3.26]
5:        phi     2.29  0.20     [1.91, 2.69]
6:        tau     6.75  0.08     [6.58, 6.91]
7:      delta     2.47  0.08      [2.3, 2.62]
--->

In our sample, the baseline RRi controlled by $\alpha$ was 876.9 ms (CI~95%~[845.5, 906.6]), with an exercise induced drop ($\beta$) of 386.6 (CI~95%~[351.0, 420.3]). The recovery proportion ($c$ parameter) relative to $\beta$ was 0.9 (CI~95%~[0.86, 0.93]).

In relation to the rate parameters, controlling the speed of transition between exercise and recovery phases, we observed a rate of exercise-induced decay ($\lambda$) of 2.96 min^-1^ (CI~95%~[2.63, 3.26]) and a post exercise recovery rate ($\phi$) of 2.29 min^-1^ (CI~95%~[1.91, 2.69]).

For time parameters controlling the timing of the aforementioned RRi kinetics, we observed that the time of the exercise-induced ($\tau$) drop aligned at 6.75 (CI~95%~[6.58, 6.91]) and a duration of the exertion-related RRi depression ($\delta$) was prolonged for 2.47 (CI~95%~[2.3, 2.62]).

@@

The overall exercise-induced RRi trajectory, previously described in the form of the parameters controlling the shape of the cardiac autonomic response, is transiently modified in the face of different quantities of B cell phenotypes. The specific morphological changes of the RRi-vs-time curve are dependent on the independent variation of both CD21 and CD11c B cell markers. The main immune-autonomic signatures derived by different B cell expression profiles can be seen in [Figure 3](#fig-3).

![](../figures/hand-designed/rri-signatures.pdf){width=100%}

[**Figure 3**]{#fig-3}. Exercise-induced RRi signatures dependent on B cell counts with CD21 and CD11c expressions. Each RRi-vs-time curve correspond to the simulation of RRi dynamics based on the predicted model parameters for the corresponding combination of B cell phenotypes (of CD21 and CD11c) below and above two standard deviations, while keeping the remaining cell markers counts constant at their mean. Depicted effects are adjusted for confounding variables.

@@

```{r}
#| eval: false
#| echo: false
report_model(model, variable = "cd21_p_cd11c_m")
```

Individuals that have more B cells with CD21, but not CD11c markers, seems to have more than 80% probability of having a less pronounced decrease in RRi (ES on $\beta$ = -0.45, CI~95%~[-1.29, 0.32], pd = 86%, ps = 80.2%), as well as a more steep exercise-induced RRi drop (ES on $\lambda$ = 0.43, CI~95%~[-0.41, 1.29], pd = 83.7%, ps = 77.3%) and recovery (ES on $\phi$ = 0.49, CI~95%~[-0.36, 1.34], pd = 87.8%, ps = 82.2%).

@@

```{r}
#| eval: false
#| echo: false
report_model(model, variable = "cd21_p_cd11c_p")
```

For individuals with more expression of both CD21 and CD11c (i.e., double positive), we observed a higher RRi at rest (ES on $\alpha$ = 0.52, CI~95%~[0.18, 0.86], pd = 99.8%, ps = 98.9%), more pronounced exercise-induced RRi decrease (ES on $\beta$ = 0.4, CI~95%~[0.06, 0.75], pd = 98.8%, ps = 95.6%), less post-exercise RRi recovery (ES on $c$ = -0.28, CI~95%~[-0.61, 0.06], pd = 95.4%, ps = 85.7%) and higher duration of RRi depression till recovery kinetics begin relative to RRi drop (ES on $\delta$ = 0.3, CI~95%~[-0.08, 0.67], pd = 94.3%, ps = 85.7%).

@@

```{r}
#| eval: false
#| echo: false
report_model(model, variable = "cd21_m_cd11c_m")
```

On the other side, individuals with higher counts of B cells absent of CD21 or CD11c markers (i.e., double negative) experienced lower baseline RRi (ES on $\alpha$ = -0.29, CI~95%~[-0.67, 0.1], pd = 92.9%, ps = 83.1%), less exercise-induced RRi decrease (ES on $\beta$ = -0.41, CI~95%~[-0.8, -0.03], pd = 98.2%, ps = 94.7%), better post-exercise recovery (ES on $c$ = 0.31, CI~95%~[-0.05, 0.69], pd = 94.9%, ps = 86.6%), steeper RRi drop (ES on $\lambda$ = 0.4, CI~95%~[-0.02, 0.81], pd = 97.2%, ps = 92.2%) and early initiation of cardiac autonomic kinematics (ES on $\tau$= -0.38, CI~95%~[-0.77, 0.05], pd = 96.3%, ps = 90.7%).

@@

```{r}
#| eval: false
#| echo: false
report_model(model, variable = "cd21_m_cd11c_p")
```

For B cells with more CD11c expression, but not CD21, we observed that greater counts of this cell phenotype were associated 73% probability of having a significant slower RRi drop initiated by exercise (ES on $\lambda$ = -0.21, CI~95%~[-0.58, 0.15], pd = 88%, ps = 73%) and close to 90% probability of observing a significant delay in exercise-induced cardiac autonomic kinematics (ES on $\tau$ = 0.32, CI~95%~[-0.04, 0.69], pd = 96%, ps = 89%).

The specific effect of each specific B cell phenotype on the different aspects of RRi dynamics can be seen in [Figure 4](#fig-4).

![](../figures/fig_params_vs_immune_adjusted.pdf){width=100%}
